---
title: "Analisa data statistik pendidikan dasar SD/SMP/SMA/SMK di Indonesia dengan bahasa pemrograman R"
author: "Hendy Irawan"
date: "June 29, 2015"
output:
  html_document:
    self_contained: no
---

**(Statistical data analysis of Indonesian basic education with R programming language)**

## Requirements

```{r eval=FALSE}
install.packages('dplyr')
install.packages('ggplot2')
```

If you just install `xlsx`/`rJava` in Linux you'll get this error:

```
Error : .onLoad failed in loadNamespace() for 'rJava', details:
  call: dyn.load(file, DLLpath = DLLpath, ...)
  error: unable to load shared object '/home/ceefour/R/x86_64-unknown-linux-gnu-library/3.1/rJava/libs/rJava.so':
  libjvm.so: cannot open shared object file: No such file or directory
Error: loading failed
```

[For Linux(Ubuntu) users](http://stackoverflow.com/a/25932828/122441): If you have oracle-java (7/8) installed. It'll be at this location `/usr/lib/jvm` and sudo access is required.

Create the file `/etc/ld.so.conf.d/java.conf` with the following entries:

    /usr/lib/jvm/java-8-oracle/jre/lib/amd64
    /usr/lib/jvm/java-8-oracle/jre/lib/amd64/server

(Replace java-8-oracle with java-7-oracle depending on your java version)

Then:

    sudo ldconfig

Restart RStudio and then install the `xslx/rJava` package.

```{r eval=FALSE}
install.packages('xlsx')
```

Unfortunately I still get error :( (probably due to http://stackoverflow.com/a/21424300/122441)

```
> un_smp <- read.xlsx('un-smp-bandung-2015.xlsx', 1)
Error in .jcall("RJavaTools", "Ljava/lang/Object;", "invokeMethod", cl,  : 
  org.apache.poi.POIXMLException: java.lang.reflect.InvocationTargetException
> un_smp <- read.xlsx2('un-smp-bandung-2015.xlsx', 1)
Error in .jcall("RJavaTools", "Ljava/lang/Object;", "invokeMethod", cl,  : 
  org.apache.poi.POIXMLException: java.lang.reflect.InvocationTargetException
```

## Analysis

```{r}
library(dplyr)
library(pander)
library(xlsx)
un_smp <- read.xlsx2('un-smp-bandung-2015.xlsx', 1)
```
